Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-07-05 edubots_surveys.Rnw
Updated the introduction section
Added the reviewed paragraphs to the introduction section.

% TODO: introduction
The introduction of new technologies in the classroom involves a combination of teacher training, student enthusiasm and possibly acquiring new skills by both. Some technologies are readily adapted, but others require longer time for full adoption. In most cases, the magnitude of the uptake of the new technology impinges on the collaboration of all the parties using it. Hence, examining the users’ attitudes and the opinions is an important first step in boosting this collaboration and reaping the benefits of the technology.  Recently, a surge in the use of synchronous instant messaging tools in higher education such as WhatsApp, Telegram and Facebook Messenger took place. Some messaging applications have built in chatbots that are able to support a synchronous conversation between the different parties \cite{differ.chat}, while other chatbots are developed as standalone systems and can possibly be attached to a messaging application using an API (Application Program Interface).

In the framework of the EduBots KA2 European project\footnote{https://www.edubots.eu}, we are mainly concerned with the creation of best practices in the use of chatbots in higher education classrooms. A \textit{chatbot} is a program, sometimes based on Artificial Intelligence techniques, that allows a computer to communicate in a similar way as a human does \cite{Gong2008}, using natural language. Indeed, most of them would be, apparently, close to passing a classic Turing test \cite{moor2003turing}, since they are able to answer almost any speaker's question fluently, and even ask other questions to the human. However, in most cases a simple messaging-application attached bot will help the students (and teachers); so neither computational intelligence nor comprehension of natural languages are requisites, the only one being the fact that they can be accessed from a synchronous conversational application such as the common instant messaging applications (e.g. Telegram, WhatsApp) used nowadays.
% Antonio - This paragraph needs a conclusion. Why do we say (essentially) that chatbots are not needed and just a simple messaging bot is enough?
% JJ - feel free to add one 
% Antonio - The thing is I don't catch why are we saying that, since it is supossed that we should 'defend' the use of chatbots (from the project perspective, at least). Or if we don't want to defend them, at least not saying that they are useless most of times. XD
% Moreover, considering that in the following paragraphs we are trying to 'sell' them.
A chatbot is a program, sometimes developed using Artificial Intelligence techniques, that allows a computer to communicate in a similar way as humans do \cite{Gong2008}, using natural language. Indeed, most of them would be, apparently, close to passing a classic Turing test \cite{ moor2003turing }, since they are able to answer almost any question fluently, and even pose questions in their conversations. The recent advancement in this discipline has played an important role in many fields especially in education and online tutoring \cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery of personalized learning by adapting to students’ pace of learning and providing customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experiences as well as improving individual attention \cite{agarwal2020}.
The technology behind chatbots \cite{Gong2008,bradesko2012} has become an usual element in our everyday life, having played an important role in the development of many fields, including education and online tutoring \cite{clarizia2018,Smutny2020}, which is what we are mainly concerned with. These automatic systems ease personalized learning, 
adapting individuals' pace of learning, who get customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experience as well as individual attention \cite{agarwal2020}.
Given the potential benefits that chatbots could bring to the classroom, the focus of the EDUBOTS KA2 European project [1], the sponsor of this study, is to gather, analyse and disseminate best practice of the use of chatbots in higher education institutions.  This study involves two phases of data collection. The first phase aimed to collect feedback from students about their actual needs \cite{MoraChatbots2021}.  Hence, the objective was to collect information about students’ preferences of tools/applications for chatting and messaging, how they use them in educational context, who do they like to be with in the class messaging groups, and their expectations of the chatbots in terms of assisting them during their learning process. Therefore, two surveys for bachelor and master degree students at the University of Granada (Spain) were developed and answers from more than 250 students were collected.   The key findings of the students surveys reveal that they prefer to use messaging applications, such as Telegram or WhatsApp as they are familiar to them. Students also expressed interest in chatbots that can assist them with organizing their course schedules, help them access their assignment grades and facilitate searching for course’s resources. 
The results of the first phase were fed into the second phase of the study, which focussed on teachers’ opinions about using chatbots and messaging applications in classes.  Another survey was developed for this purpose. Responses from 300 teachers were collected and analysed.  This study focuses on discussing the implications and the impact of the development of future technology-enhanced tools and educational institutions’ policies by finding answers to the following research questions: 
• RQ1 - Are teachers already using messaging apps in their classes? 
• RQ2 - Which kind of chatbots would teachers find useful in their classes? 
• RQ3 - Which kind of interaction do teachers prefer with their students? 
The remainder of the paper is organized as follows: Next is an overview of what current research has found about the use of messaging applications, including chatbots, in the classroom. The methodology used in the survey is presented in Section 3, and the results of the survey are presented next in Section 4. Finally, we discuss these results and conclude with a series of recommendations in the section that closes the paper.

In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, students suffer from stress when they have to ask a question in front of the entire class. This is why many students prefer to remain silent and ask their question later via email. 
However, this means a waste of time and a delay of hours or even days to get the teacher's answer; moreover, it does not scale well with the size of the class.
Chatbot technology has a potential to fill the gap between teacher and students, helping them to solve questions and carry out a dynamic and autonomous learning \cite{Griol2014,kim2020}.
In addition, using an automated system such as a chatbot, allows the teacher to detect the most demanded topics, and on which he or she should consequently place more emphasis in class.

Chatbots are obviously associated to messaging applications, and in particular, direct messaging applications, since they are the only type of applications that are able to support a synchronous conversation between the two parties. This is why in many cases chatbots couple messaging and the chatbot technology itself \cite{differ.chat}, while in others, you can create chatbots in any number of technologies and then attach them to a messaging application using their API (Application Program Interface).

Given the potential advantages that chatbots could bring to the classroom, as a previous step, we created two surveys for bachelor and master degree students at the University of Granada (Spain), which were answered by more than 250 students. The obtained results were analysed with the object of identifying students' actual needs, in \cite{MoraChatbots2021}. The survey asked about their preferences on tools/applications for chatting and messaging, how do they use them in the educational context, who do they like to be with in the class messaging groups, and what do they expect bots and chatbots to do to aid them during their learning process. We concluded that students prefer to use known whatever messaging application, such as Telegram or WhatsApp, they are already using. They also expected chatbots to help them with their scheduling, to provide assignment grades and to search for class material. Only one of the applications mentioned, having them work with frequently asked questions, might actually require some understanding of natural language, although it can be formulated in such a way that is not needed.
 In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, some students experience stress because of their need to ask a question in front of the entire class. This is why many students refrain to silence in the classroom and tend to contact their teachers later via email. Consequently, this can lead to not only a delay in obtaining answers to their questions but also can cause a significant increase in their teachers’ workload, especially when factoring in the disproportion between the class size and the number of teachers. Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding. 
% Maybe move it to a different place later
In the present work we aim to complement the aforementioned study with an analysis of teacher attitudes; teachers are the ones that actually would need to use, 
% Olia-I believe we should avoid refereeing to “attitudes”. Also here, it is not mentioned that we explore their views towards the use of chat tools/ messaging apps. 
% Antonio - 'deploy' instead implement? Since in Computer Sciences implement and program are synonyms...
% JJ - deployment would be done by the university or whoever publishes the system. Deployment and implementation are two different things, anyway
% Antonio - Ok, but what I said is that 'implement' is a synonym of 'program' in this scope, and we write both in the same sentence as different things.
% And deploy does not make sense in this context. I will simply say use - JJ
configure and maybe program chatbots. Thus, we have performed a survey among 300 teachers asking about their attitudes towards the use of bots or chatbots in their classes. We analyse their responses, since we are interested in discussing implications and the impact for the development of future technology-enhanced tools and the Education institutions policies regarding the following research questions:
\begin{itemize}
\item{ RQ1 - Are teachers already using messaging apps in their classes? } %This was the second survey, not the  first survey. I wonder if maybe we could invert the presentation order - JJ
\item{ RQ2 - Which kind of chatbots would teachers find useful in their classes? }
\item{ RQ3 - Which kind of interaction do teachers prefer with their students? }
\item{ RQ4 - What kind of interaction media features do teachers value the most? }

@@

% Antonio - I include here the description of interaction features. Not sure if it is the best place for it. Move wherever it better fits.
% Olia- in the text below, we refer to "communucation tools", we should be consisten with the terms we use. Should we refere to communication tools? Messaging apps? Chat tools? 
Specifically, this survey asked the teachers, in addition to general information (such as gender, age, experience, etc), what are their feelings and preferences with regard to the use of communication tools with students, and even using bots or chatbots in the classroom.
In addition the survey asked about the most relevant \textit{Interaction media features} for the teachers with respect to the potential tool to use in class. These features are:
\begin{itemize}
\item \textit{Analytics}: the tool allows to extract data about students performance or development during the academic course on the teacher's subject.
\item \textit{Connectivity}: it is possible to link or connect the tool with other applications.
\item \textit{Familiarity}: it is something already known or previously used, so no need to learn something new.
\item \textit{Hidden Phone}: it is possible to use the tool without showing my phone number, normally using an alias or an own account independent from my physical telephone.
\item \textit{Horizontally}: all the actors are at the same level (for instance, teacher and students). There is not a hierarchy.
\item \textit{Official formation}: the university offers some training courses or talks about the tool (mainly for teachers). 
\item \textit{Pluggability}: the tool offers different pluggins to the teacher, so he/she can deploy his/her own implemented modules.
\item \textit{Sustainability}: teacher should not be worried about the tool maintenance.
\item \textit{Unidirectionality}: only the teacher can publish information in the channel, i.e. students are not allowed to ask or answer.
\item \textit{Officiality}: the tool has been granted by the university where the teacher works. This means normally that there is also technical support by university's staff.
\item \textit{Synchronous communication}: the interactions between teacher and students take place in real time, i.e. it is not a delayed communication scheme such as e-mail.
\end{itemize}

@@

% Olia - TODO: check for repetitions in the following paragraph that you added. thank you.
The first column in Fig. \ref{figure:postcovid_chisq} shows the Pearson's residuals of the test for these three variables. Residuals show the difference between the observed and the expected values, with large residuals showing that variables are not truly independent. In this case, blue shows positive contribution and red negative contribution to the test, and saturation shows how large is this contribution in contrast to the expected value by chance. Tables on the right side show adjusted standardized residuals that according to \cite{agresti2013categorical}, if greater than +/-2 for cases with few cells, indicate lack of fit of $H_0$ (bolded in our table). Also, the table shows the percentage of contribution to the test of each cell.

@@

Answer to these questions are plotted in Figure \ref{figure:chatbots_types} grouped by gender, sector and experience. As it was a multiple choice question, the counts are over the total number of people that answered. One of the most interesting questions that can arise from the plots, and the should require further study, is: Are teachers really aware of the possibilities of chatbots in the classroom? The reason to ask ourselves this question is due to the fact of the small number of responses for other uses in comparison with the other approaches proposed. Another way to look about it is that, unless you provide the specific instructions on how to use a tool, the teaching staff is already too busy or burn out to think about innovation.

@@

%Olia - move figure 7 elsewhere, not at the beginning of this section (RQ3). 
Since chatbots are intended to mediate or help in this interaction, it is essential to understand the kind of interaction teachers prefer. Chatbots should address those modes, and not others. In order to find out these modes, and thus answering RQ3, teachers were provided with a list of different kind of interactions that may take place among students and between students and the teacher, with the use of messaging apps. The findings are summarized in Table \ref{tab:kindinteractions}.

@@

 \item Keep it Simple. Media interaction requirements are simple and only a 2.4\% is requesting more features. This could be a sign of technological burn out produced by the previous and current courses where, due to the pandemic, the use of computer has increased significantly. 
\end{itemize}
%Olia: be consistent in the terms being used, we refer to teachers, then to professors. I would replace the word "professors" above with the word "teachers".

@@

The key findings of this study shed the light on educators’ preference to use messaging applications that their institutions support.  Technology adoption literature often focus on users’ perceptions of the technology’s usefulness and ease of use as important prerequisites for successful adoption and utilization. Nevertheless, in higher education, institutions’ role in integrating these tools to their educational systems can improve the uptake of these applications and shape the social and educational experiences of their students.  This integration, to be fruitful, institutions have to ensure that these messaging applications not only are GDPR compliant to keep students’ data secure but also should provide IT support to all stakeholders who use the applications .  
Moreover, this study clarifies the differences between the teachers’ perspectives and/or intentions and that of their students when using the messaging applications in higher education.  Teachers tend to adopt technologies supported by their institutions.  This could be to ensure that their universities can oversee their efforts in supporting their students during the learning process.  Another reason could be their familiarity with the technology provided by their institutions. On the other hand, students use the non-institution messaging applications to form informal discussion groups with their peers.  It is worth noting that peer support and collaboration is inseparable from learning \cite{timmis2012constant} and correlates positively with higher retention in higher education \cite{o2014front}.  Therefore, both perspectives are complementary and play different roles in promoting the learning process.
%This section is repeated
%\section*{Acknowledgements}

>>>>

% TODO: introduction

The introduction of new technologies in the classroom involves a combination of teacher training, student enthusiasm and possibly acquiring new skills by both. Some technologies are readily adapted, but others require longer time for full adoption. In most cases, the magnitude of the uptake of the new technology impinges on the collaboration of all the parties using it. Hence, examining the users’ attitudes and the opinions is an important first step in boosting this collaboration and reaping the benefits of the technology.  Recently, a surge in the use of synchronous instant messaging tools in higher education such as WhatsApp, Telegram and Facebook Messenger took place. Some messaging applications have built in chatbots that are able to support a synchronous conversation between the different parties \cite{differ.chat}, while other chatbots are developed as standalone systems and can possibly be attached to a messaging application using an API (Application Program Interface).

A chatbot is a program, sometimes developed using Artificial Intelligence techniques, that allows a computer to communicate in a similar way as humans do \cite{Gong2008}, using natural language. Indeed, most of them would be, apparently, close to passing a classic Turing test \cite{ moor2003turing }, since they are able to answer almost any question fluently, and even pose questions in their conversations. The recent advancement in this discipline has played an important role in many fields especially in education and online tutoring \cite{clarizia2018,Smutny2020}. These automatic systems facilitate the delivery of personalized learning by adapting to students’ pace of learning and providing customized online tutoring outside the classroom. Thus, chatbots can significantly contribute to providing interactive learning experiences as well as improving individual attention \cite{agarwal2020}.
 In this sense, chatbot technology offers a great opportunity for the improvement of tutoring systems \cite{Daniel2016,agarwal2020}, as not all students are comfortable with face-to-face tutoring with the instructor. In fact, in many cases, some students experience stress because of their need to ask a question in front of the entire class. This is why many students refrain to silence in the classroom and tend to contact their teachers later via email. Consequently, this can lead to not only a delay in obtaining answers to their questions but also can cause a significant increase in their teachers’ workload, especially when factoring in the disproportion between the class size and the number of teachers. Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding. 
Given the potential benefits that chatbots could bring to the classroom, the focus of the EDUBOTS KA2 European project \footnote{https://www.edubots.eu}, the sponsor of this study, is to gather, analyse and disseminate best practice of the use of chatbots in higher education institutions.  This study involves two phases of data collection. The first phase aimed to collect feedback from students about their actual needs \cite{MoraChatbots2021}.  Hence, the objective was to collect information about students’ preferences of tools/applications for chatting and messaging, how they use them in educational context, who do they like to be with in the class messaging groups, and their expectations of the chatbots in terms of assisting them during their learning process. Therefore, two surveys for bachelor and master degree students at the University of Granada (Spain) were developed and answers from more than 250 students were collected.   The key findings of the students surveys reveal that they prefer to use messaging applications, such as Telegram or WhatsApp as they are familiar to them. Students also expressed interest in chatbots that can assist them with organizing their course schedules, help them access their assignment grades and facilitate searching for course’s resources. 
The results of the first phase were fed into the second phase of the study, which focussed on teachers’ opinions about using chatbots and messaging applications in classes.  Another survey was developed for this purpose. Responses from 300 teachers were collected and analysed.  This study focuses on discussing the implications and the impact of the development of future technology-enhanced tools and educational institutions’ policies by finding answers to the following research questions:

% Antonio - This paragraph needs a conclusion. Why do we say (essentially) that chatbots are not needed and just a simple messaging bot is enough?
% JJ - feel free to add one 
% Antonio - The thing is I don't catch why are we saying that, since it is supossed that we should 'defend' the use of chatbots (from the project perspective, at least). Or if we don't want to defend them, at least not saying that they are useless most of times. XD
% Moreover, considering that in the following paragraphs we are trying to 'sell' them.

% Maybe move it to a different place later
In the present work we aim to complement the aforementioned study with an analysis of teacher attitudes; teachers are the ones that actually would need to use, 
% Antonio - 'deploy' instead implement? Since in Computer Sciences implement and program are synonyms...
% JJ - deployment would be done by the university or whoever publishes the system. Deployment and implementation are two different things, anyway
% Antonio - Ok, but what I said is that 'implement' is a synonym of 'program' in this scope, and we write both in the same sentence as different things.
% And deploy does not make sense in this context. I will simply say use - JJ
configure and maybe program chatbots. Thus, we have performed a survey among 300 teachers asking about their attitudes towards the use of bots or chatbots in their classes. We analyse their responses, since we are interested in discussing implications and the impact for the development of future technology-enhanced tools and the Education institutions policies regarding the following research questions:
\begin{itemize}
\item{ RQ1 - Are teachers already using messaging apps in their classes? } %This was the second survey, not the  first survey. I wonder if maybe we could invert the presentation order - JJ
\item{ RQ2 - Which kind of chatbots would teachers find useful in their classes? }
\item{ RQ3 - Which kind of interaction do teachers prefer with their students? }
\item{ RQ4 - What kind of interaction media features do teachers value the most? }

@@

% Antonio - I include here the description of interaction features. Not sure if it is the best place for it. Move wherever it better fits.
Specifically, this survey asked the teachers, in addition to general information (such as gender, age, experience, etc), what are their feelings and preferences with regard to the use of communication tools with students, and even using bots or chatbots in the classroom.
In addition the survey asked about the most relevant \textit{Interaction media features} for the teachers with respect to the potential tool to use in class. These features are:
\begin{itemize}
\item \textit{Analytics}: the tool allows to extract data about students performance or development during the academic course on the teacher's subject.
\item \textit{Connectivity}: it is possible to link or connect the tool with other applications.
\item \textit{Familiarity}: it is something already known or previously used, so no need to learn something new.
\item \textit{Hidden Phone}: it is possible to use the tool without showing my phone number, normally using an alias or an own account independent from my physical telephone.
\item \textit{Horizontally}: all the actors are at the same level (for instance, teacher and students). There is not a hierarchy.
\item \textit{Official formation}: the university offers some training courses or talks about the tool (mainly for teachers). 
\item \textit{Pluggability}: the tool offers different pluggins to the teacher, so he/she can deploy his/her own implemented modules.
\item \textit{Sustainability}: teacher should not be worried about the tool maintenance.
\item \textit{Unidirectionality}: only the teacher can publish information in the channel, i.e. students are not allowed to ask or answer.
\item \textit{Officiality}: the tool has been granted by the university where the teacher works. This means normally that there is also technical support by university's staff.
\item \textit{Synchrony}: it is possible to publish/receive the data or information in several devices and all of them will contain or show the same.
\end{itemize}

@@

% Olia - TODO: check for repetitions in the following paragraph that you added. thank you.
The descriptive analysis in this case is done using a Chi-square test. For the different variables considered, some statistical significance was found for Gender, Sector, and academic Discipline. The values found with the test are: $p-value=0.0034$ for Gender, $p-value=0.0199$ for Sector, and $p-value=0.0211$ for Discipline.
The first column in Fig. \ref{figure:postcovid_chisq} shows the Pearson's residuals of the test for these three variables. Residuals show the difference between the observed and the expected values, with large residuals showing that variables are not truly independent. In this case, blue shows positive contribution and red negative contribution to the test, and saturation shows how large is this contribution in contrast to the expected value by chance. Tables on the right side show adjusted standardized residuals that according to \cite{agresti2013categorical}, if greater than +/-2 for cases with few cells, indicate lack of fit of $H_0$ (bolded in our table). Also, the table shows the percentage of contribution to the test of each cell.

@@

Answer to these questions are plotted in Figure \ref{figure:chatbots_types} grouped by gender, sector and experience. As it was a multiple choice question, the counts are over the total number of people that answered. One of the most interesting questions that can arise from the plots, and the should require further study, is: Are our professors really aware of the possibilities of chatbots in the classroom? The reason to ask ourselves this question is due to the fact of the small number of responses for other uses in comparison with the other approaches proposed. Another way to loook about it is that, unless you provide the specific instructions on how to use a tool, the teaching staff is already too busy or burn out to think about innovation.

@@

Since chatbots are intended to mediate or help in this interaction, it is essential to understand the kind of interaction teachers prefer. Chatbots should address those modes, and not others. In order to find out these modes, and thus answering RQ3, teachers were provided with a list of different kind of interactions that may take place among students and between students and the teacher, with the use of messaging apps. The findings are summarized in Table \ref{tab:kindinteractions}.

@@

 \item Keep it Simple. Media interaction requirements are simple and only a 2.4\% is requesting more features. This could be a sign of technological burn out produced by the previous and current courses where, due to the pandemic, the use of computer has increased significantly. 
\end{itemize}

@@

The key findings of this study shed the light on educators’ preference to use messaging applications that their institutions support.  Technology adoption literature often focus on users’ perceptions of the technology’s usefulness and ease of use as important prerequisites for successful adoption and utilization. Nevertheless, in higher education, institutions’ role in integrating these tools to their educational systems can improve the uptake of these applications and shape the social and educational experiences of their students.  This integration, to be fruitful, institutions have to ensure that these messaging applications not only are GDPR compliant to keep students’ data secure but also should provide IT support to all stakeholders who use the applications .

%This section is repeated
%\section*{Acknowledgements}
